Spike spectra for recurrences

### 2022-11-01 Manuscript/inter_spike_manuscript_mdpi.tex
More questions answered of Reviewer 1 and 2. Mehrnaz' comments incorporated

\citet{Zbilut2008} pointed out that $\tau$-RR could be interpreted as analogous to the auto-correlation function $C(\tau)$ and, hence, via the Wiener-Khinchin theorem, provide an analogue,
``generalized'', power spectral density. This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR. There are clear advantages for a recurrence-derived 
power spectral density, i.e., Fourier transforming (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E), instead 
of $C(\tau)$: There are no assumptions for stationarity or sampling, when constructing a RP.
Furthermore, the correlation structures of higher dimensional spaces can be resolved in the recurrence-derived Fourier-spectrum.\\

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. This novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Section \ref{sec_tau_rr_method} explains the idea. Note that this approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. However, the more spiky the signal is, the more useful our new approach is compared to a 
FT. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

@@

\noindent Eventually, an under-determined linear system
\begin{equation}
\bfX^T \boldsymbol{\beta}=\vec{s}
\label{eq_linear_system}
\end{equation}  
has to be solved for $\boldsymbol\beta$, a $(M\times 1)$-sized vector carrying the loadings we are interested in. Along a variety of algorithms which can solve this problem, we are 
particularly interested in those solutions, which promote sparsity in $\boldsymbol\beta$, since our goal is to decompose the signal $\vec{s}$ into a minimal number of basis 
functions (for an excellent overview of the topic we refer to \citet{Brunton2019}). In this paper we either use the \textit{least absolute shrinkage and selection operator} 
(LASSO) \cite{Tibshirani1996} or a \textit{sequentially thresholded least squares} (STLS) regression \cite{Brunton2016,Brunton2019} to obtain a solution $\hat{\boldsymbol\beta}$. Finally, we group loadings 
which correspond to basis functions having the same period 
$T_\text{is}$ into $\hat{\boldsymbol\beta}_f$ and obtain the 
\textit{inter-spike spectrum} by simply plotting $\hat{\boldsymbol\beta}_f$ as a function of the frequency $f=T_\text{is}^{-1}$, with $T_\text{is}=\Delta t, 2\Delta t, \dots, (\ceil*{N/2}+1) \Delta t$ 
(Fig.~\ref{fig_tau_rr_dirac_comb}C, F).

@@

When introducing a sparsity regularization we give up the ability to fully reconstruct the signal from the decomposition in favor of identifying the most important periodicities more clearly. This choice is mediated by a regression regularization parameter $\alpha$. Only for $\alpha = 0$ do we obtain a solution to Eq.~(\ref{eq_linear_system}) and thus a full reconstruction of the signal. Our general approach to fixing $\alpha$ is to require that we preserve most of the original siginal in the sense of matching a given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and the reconstruction.

@@

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_laskar_spectra}
 \caption{\textbf{A} Evolutionary Fourier power spectra of eccentricity time series. \textbf{B} Inter-spike spectrogram of the $\tau$-recurrence rate of the eccentricity time 
 series and \textit{C} its Fourier spectrogram. Horizontal black dashed lines highlight the analytically expected orbital periods of $405$, $124$ and $95$ kyrs. 
 For comparability, in all cases the spectra aligned in the columns of the shown plots are normalized to probabilities (sum of unity for 
 each power spectrum). For further computational details, please refer to the main text.}  
\label{fig_laskar_spectra}
\end{figure}

@@

We have successfully used the idea of transforming the $\tau$-RR for the detection of bifurcations in the Logistic map. By constructing appropriate surrogates of the inter-spike spectra, 
and thus a null model, the number of significant peaks in the inter-spike spectrum correlated well with the positive Lyapunov exponent. This measure was also able to resolve 
period-doubling bifurcations. However, we have to admit this way the detection of a bifurcation is only possible when the additional period(s) is not an integer multiple of the former period(s). 
This behavior is described in the application to the R\"ossler system, where we explicitly showed the different inter-spike power spectra for period-2, period-3, and chaotic dynamics.
Further development might potentially incorporate the mutual distance of peaks in the spectrum for a better correlation to the Lyapunov exponent. The inter-spike spectra of power grid 
frequency data illustrate that our proposed method may serve as a valuable source of information in addition to a standard Fourier analysis. And last but not least, we showed that our approach 
really incorporates frequencies, which are apparent in the Fourier spectrum of the signal, by applying it to analytically derived eccentricity data, where the dominant frequencies are well 
known.

@@

A novel type of power spectral density, the \textit{inter-spike spectrum}, has been proposed. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution. We chose LASSO or a sequentially 
thresholded least squares regression STLS in this letter. Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter-spike spectrum is obtained. 
Although this novel power spectrum is superior to an ordinary FFT-based power spectrum when the signal has a spike-train-like appearance, the authors suggest that this method should 
be considered 
as an additional source of information but not as a substitute for ordinary Fourier analysis. 
Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter-spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

>>>>

\citet{Zbilut2008} pointed out that $\tau$-RR could be interpreted as analogous to the auto-correlation function $C(\tau)$ and, hence, via the Wiener-Khinchin theorem, provide an analogue,
``generalized'', spectral density. This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR. There are clear advantages for a recurrence-derived spectral density, 
i.e., Fourier transforming (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E), instead 
of $C(\tau)$: There are no assumptions for stationarity or sampling, when constructing a RP.
Furthermore, since an RP can represent high-dimensional dynamics and its $\tau$-RR serves as a plug-in for $C(\tau)$, the correlation structures of higher dimensional spaces can be read from 
the recurrence-derived Fourier-spectrum.\\

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. This novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies in such a way that harmonics of these frequencies are also excited (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Section \ref{sec_tau_rr_method} explains the idea. Note that this approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. However, the more spiky the signal is, the more useful our new approach is compared to a 
FT. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the min-max-normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

@@

\noindent Eventually, an under-determined linear system
\begin{equation}
\bfX^T \boldsymbol{\beta}=\vec{s}
\label{eq_linear_system}
\end{equation}  
has to be solved for $\boldsymbol\beta$, a $(M\times 1)$-sized vector carrying the loadings we are interested in. Along a variety of algorithms which can solve this problem, we are 
particularly interested in those solutions, which promote sparsity in $\boldsymbol\beta$, since our goal is to decompose the signal $\vec{s}$ into a minimal number of basis 
functions (for an excellent overview of the topic we refer to \citet{Brunton2019}). In this paper we either use the \textit{least absolute shrinkage and selection operator} 
(LASSO) \cite{Tibshirani1996} or a \textit{sequentially thresholded least squares} (STLS) regression \cite{Brunton2016,Brunton2019} to obtain a solution $\hat{\boldsymbol\beta}$. Any other 
sparse regression technique could be used. Finally, we group loadings 
which correspond to basis functions having the same period 
$T_\text{is}$ into $\hat{\boldsymbol\beta}_f$ and obtain the 
\textit{inter-spike spectrum} by simply plotting $\hat{\boldsymbol\beta}_f$ as a function of the frequency $f=T_\text{is}^{-1}$, with $T_\text{is}=\Delta t, 2\Delta t, \dots, (\ceil*{N/2}+1) \Delta t$ 
(Fig.~\ref{fig_tau_rr_dirac_comb}C, F).

@@

When introducing a sparsity regularization we give up the ability to fully reconstruct the signal from the decomposition in favor of identifying the most important periodicities more clearly. This choice is mediated by a regression regularization parameter $\alpha$. Only for $\alpha = 0$ do we obtain a solution to Eq.~(\ref{eq_linear_system}) and thus a full reconstruction of the signal. Our general approach to fixing $\alpha$ is to require that we preserve most of the original signal in the sense of matching a given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and the reconstruction. The chosen regression method (in our case, LASSO or STLS) will determine how sensitive the obtained inter-spike spectrum is with respect to the regularization given by a 
selected $\rho$ (Fig.~\ref{fig_regularization}). Regarding this, in most of our applications LASSO yields more robust results, while STLS is usually faster to compute. 
We will discuss this in more detail in Section 4. Section \ref{sec_tau_rr_discussion}.

@@

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_laskar_spectra}
 \caption{\textbf{A} Evolutionary Fourier power spectra of eccentricity time series. \textbf{B} Inter-spike spectrogram of the $\tau$-recurrence rate of the eccentricity time 
 series and \textbf{C} its Fourier spectrogram. Horizontal black dashed lines highlight the analytically expected orbital periods of $405$, $124$ and $95$ kyrs. 
 For comparability, in all cases the spectra aligned in the columns of the shown plots are normalized to probabilities (sum of unity for 
 each power spectrum). For further computational details, please refer to the main text.}  
\label{fig_laskar_spectra}
\end{figure}

@@

We have successfully used the idea of transforming the $\tau$-RR for the detection of bifurcations in the Logistic map. By constructing appropriate surrogates of the inter-spike spectra, 
and thus a null model, the number of significant peaks in the inter-spike spectrum correlated well with the positive Lyapunov exponent. This measure was also able to resolve 
period-doubling bifurcations. However, we have to admit this way the detection of a bifurcation is only possible when the additional period(s) is not an integer multiple of the former period(s). 
This behavior is described in the application to the R\"ossler system, where we explicitly showed the different inter-spike spectra for period-2, period-3, and chaotic dynamics.
Further development might potentially incorporate the mutual distance of peaks in the spectrum for a better correlation to the Lyapunov exponent. The inter-spike spectra of power grid 
frequency data illustrate that our proposed method may serve as a valuable source of information in addition to a standard Fourier analysis. And last but not least, we showed that our approach 
really incorporates frequencies, which are apparent in the Fourier spectrum of the signal, by applying it to analytically derived eccentricity data, where the dominant frequencies are well 
known.

@@

A novel decomposition technique is proposed that yields the \textit{inter-spike spectrum}. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution. We chose LASSO or a sequentially 
thresholded least squares regression STLS in this letter. Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter-spike spectrum is obtained. 
Although this novel spectrum is superior to an ordinary FFT-based power spectrum when the signal has a spike-train-like appearance, the authors suggest that this method should 
be considered 
as an additional source of information but not as a substitute for ordinary Fourier analysis. 
Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter-spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.
